Barriers to Green Living: Insights from 2018 and 2021 Toronto Climate Perception Surveys

### 2024-12-04 paper/climate_change.qmd
working on last minute changes

Climate change is one of the most urgent challenges we face today. From rising temperatures to extreme weather events, the planet is already feeling the consequences of human-driven environmental change. Yet, despite widespread awareness of these risks, many people still struggle to adopt behaviors that could help mitigate climate damage. This gap between knowledge and action is a central puzzle that governments, communities, and organizations are trying to solve. If we are to tackle this crisis effectively, we need to better understand what stops people from acting, and how we can motivate change.

This paper examines how specific factors—such as age, education, and self-reported knowledge of climate change—shape people’s willingness to engage in climate-friendly actions. By analyzing data from a 2018 survey, we explore the connection between demographic characteristics and behaviors like reducing energy consumption, using greener transportation, and lowering waste. The survey also delves into what prevents people from adopting these behaviors, uncovering obstacles like cost, convenience, and skepticism about the power of individual actions. With this analysis, we aim to uncover how these factors interact and what they mean for efforts to encourage sustainable choices.

The heart of this study is to measure how age, education, and knowledge of climate change influence the likelihood of adopting certain behaviors. We found that younger individuals and those with higher education levels are more likely to engage in actions like reducing car usage and adopting plant-based diets. However, older individuals are more inclined to take steps like cutting down on household energy use. Beyond these demographic trends, we also discovered that barriers such as financial cost and doubts about the effectiveness of individual efforts often hold people back from taking action. These findings suggest that a “one-size-fits-all” approach to climate communication won’t work. Instead, targeted strategies that consider demographic differences are essential for driving meaningful engagement.

Understanding these patterns is crucial for crafting effective climate policies. By identifying which groups are more likely to act and which face significant barriers, we can better design interventions that speak to people’s concerns and motivations. The implications of this study go beyond simply knowing what people think about climate change—they offer a roadmap for shaping messages and policies that could shift behavior on a larger scale.

@@

```{r}
#| warning: false
#| echo: false
#| message: false
# #| label: table-one
# #| fig-cap: "Comparison of Age Distribution Between 2018 and 2021 Surveys: This table highlights the age distribution of survey respondents in 2018 and 2021. Notable differences include a higher representation of individuals aged 18–23 in 2021 compared to 2018, while the distribution across other age groups remained relatively consistent between the two years."

# Read the cleaned CSV into a data frame
# age_summary_combined <- read.csv(here("data/03-figures_data", "age_summary_combined.csv"))

# Print the cleaned table (using kable for styling)
# age_summary_combined %>%
  # kbl(caption = "Age Distribution Comparison for 2018 and 2021") %>%
  # kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

@@

# Model {#sec-model}
The goal of our modelling strategy is twofold. Firstly, we aim to predict the likelihood of individuals engaging in climate-friendly actions based on demographic variables, specifically age and education. The actions considered include investing in electric or hybrid vehicles, making home improvements (such as using LED lightbulbs), adopting meat alternatives, minimizing car use, purchasing green products, reducing hydro usage, reducing waste, sorting waste correctly, and walking or cycling shorter distances. Secondly, we aim to understand how these demographic characteristics influence individuals' likelihood of adopting these behaviors.

Here we describe the probability tree model used to investigate the relationship between likelihood and the demographic factors of age and education. Further details on background and diagnostics can be found in [Appendix -@sec-appendix].

@@

## Model Takeaways
summary table age and education in predicting likelihood of taking climate actions
the accuracy of the models prediction matches the genuine data
TODO 
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: table-five
#| fig-cap: TODO

@@

## Age and Climate-Friendly Behaviors {#sec-first-point}
The findings indicate that younger individuals are more likely to engage in climate-friendly actions that are perceived as low-cost and easier to implement, such as consuming meat alternatives and reducing car usage. This suggests that younger generations, likely due to heightened environmental awareness and exposure to climate change discussions, are more inclined to adopt behaviors that align with popular environmental movements. It also highlights how lifestyle factors and generational attitudes toward climate change can shape behavioral patterns.

## The Role of Education in Sustainable Actions {#sec-second-point}
Education significantly influences individuals' likelihood to engage in more substantial climate-friendly behaviors. Higher educational attainment is strongly associated with actions that require greater investment, such as purchasing electric vehicles or making energy-efficient home improvements. This pattern suggests that educational levels not only increase awareness but also enhance the capacity to implement larger, more financially demanding sustainable practices. The relationship between education and sustainability highlights the importance of knowledge and informed decision-making in supporting climate-friendly actions.

## The Influence of Climate Change Knowledge {#sec-third-point}
Self-reported knowledge of climate change also plays a role in the likelihood of adopting sustainable behaviors. People with greater awareness of climate change tend to be more proactive in making environmentally conscious choices. However, while knowledge is an important factor, it does not always guarantee significant behavioral change. This suggests that raising awareness must be paired with policies that reduce barriers to adopting more expensive or long-term sustainable actions, ensuring that individuals are both informed and able to make practical changes.

@@

The primary limitation of this study lies in its reliance on self-reported data, which is subject to bias. Respondents may overestimate their awareness of climate change or misreport their actions, especially when social desirability factors are involved. Additionally, the study used summary statistics for 2021, which limited the ability to explore individual-level trends and interactions. Without access to more granular data, it was impossible to assess how different socio-economic factors, such as income, directly influence individuals’ likelihood to adopt climate-friendly behaviors.

Another limitation is that the study did not explore the reasons behind individuals' reluctance to adopt certain sustainable behaviors. Investigating these reasons would provide more insight into how policies can make climate actions more accessible, affordable, and normalized for different demographic groups. Although time constraints prevented this deeper analysis, it would be an important direction for future research.

TO DO: add / talk about difficulty and problems with survey data

## Future Directions
Future research should aim to identify the specific barriers that prevent certain groups from engaging in climate-friendly behaviors. This could involve examining factors such as income, access to resources, and perceived obstacles to adopting sustainable practices. In addition, exploring how policies or interventions can make sustainable actions more accessible and affordable across different demographic groups would be valuable. Longitudinal studies could also help track how awareness of climate change influences sustained behavioral changes, and whether other external factors, such as economic incentives, contribute to these shifts.

>>>>

Climate change is one of the most urgent challenges we face today. From rising temperatures to extreme weather events, the planet is already feeling the consequences of human-driven environmental change [@IPCC2023]. Yet, despite widespread awareness of these risks, many people still struggle to adopt behaviors that could help mitigate climate damage [@Gifford2011]. This gap between knowledge and action is a central puzzle that governments, communities, and organizations are trying to solve [@Kollmuss2002]. If we are to tackle this crisis effectively, we need to better understand what stops people from acting, and how we can motivate change [@Steg2014].

This paper examines how specific factors—such as age, education, and self-reported knowledge of climate change—shape people’s willingness to engage in climate-friendly actions [@Schultz2002]. By analyzing data from a 2018 survey [@climateperceptionstudy], we explore the connection between demographic characteristics and behaviors like reducing energy consumption, using greener transportation, and lowering waste. The survey also delves into what prevents people from adopting these behaviors, uncovering obstacles like cost, convenience, and skepticism about the power of individual actions [@Gifford2011]. With this analysis, we aim to uncover how these factors interact and what they mean for efforts to encourage sustainable choices [@Stern2000].

The heart of this study is to measure how age, education, and knowledge of climate change influence the likelihood of adopting certain behaviors. We found that younger individuals and those with higher education levels are more likely to engage in actions like reducing car usage and adopting plant-based diets [@Fulton2022]. However, older individuals are more inclined to take steps like cutting down on household energy use [@Lee2019]. Beyond these demographic trends, we also discovered that barriers such as financial cost and doubts about the effectiveness of individual efforts often hold people back from taking action [@Kollmuss2002]. These findings suggest that a “one-size-fits-all” approach to climate communication won’t work. Instead, targeted strategies that consider demographic differences are essential for driving meaningful engagement [@Steg2014].

Understanding these patterns is crucial for crafting effective climate policies. By identifying which groups are more likely to act and which face significant barriers, we can better design interventions that speak to people’s concerns and motivations [@Stern2000]. The implications of this study go beyond simply knowing what people think about climate change—they offer a roadmap for shaping messages and policies that could shift behavior on a larger scale [@Fulton2022].

@@

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: table-one
#| fig-cap: "Comparison of Age Distribution Between 2018 and 2021 Surveys: This table highlights the age distribution of survey respondents in 2018 and 2021. Notable differences include a higher representation of individuals aged 18–23 in 2021 compared to 2018, while the distribution across other age groups remained relatively consistent between the two years."

# Read the cleaned CSV into a data frame
age_summary_combined <- read.csv(here("data/03-figures_data", "age_summary_combined.csv"))

# Print the cleaned table (using kable for styling)
age_summary_combined %>%
  kbl(caption = "Age Distribution Comparison for 2018 and 2021") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

@@

# Model {#sec-model}
The goal of our modelling strategy is twofold. Firstly, we aim to predict the likelihood of individuals engaging in climate-friendly actions based on demographic variables, specifically age and education. The actions considered include investing in electric or hybrid vehicles, making home improvements (such as using LED lightbulbs), adopting meat alternatives, minimizing car use, purchasing green products, reducing hydro usage, reducing waste, sorting waste correctly, and walking or cycling shorter distances. Secondly, we aim to understand how these demographic characteristics influence individuals' likelihood of adopting these behaviors.Here we describe the probability tree model used to investigate the relationship between likelihood and the demographic factors of age and education.

@@

## Model Summary Table
The table presents the accuracy of models predicting the likelihood of various climate-friendly actions based on demographic factors such as age and education. Each row represents a different climate action (e.g., home improvement, reducing hydro, minimizing car use), with the corresponding accuracy of the model's predictions. The accuracy values reflect how well the models, which use age and education as predictors, align with the actual data. An accuracy of 33 percent  for home improvement" indicates that the model correctly predicted the outcome 33% of the time for this action. These accuracy values help evaluate the effectiveness of age and education as predictors for each specific climate action.
```{r}
#| warning: false
#| echo: false
#| message: false
#| label: table-five
#| fig-cap: This table presents the accuracy of models predicting the likelihood of various climate-friendly actions based on age and education. The accuracy values reflect how closely the model predictions align with the actual observed data. Higher accuracy values indicate better model performance.

@@

## Age and Climate-Friendly Behaviors {#sec-first-point}
The findings indicate that younger individuals are more likely to engage in climate-friendly actions perceived as low-cost and easier to implement, such as consuming meat alternatives and reducing car usage. The model results show that these behaviors have relatively higher predicted likelihoods among younger participants, aligning with existing research on generational differences in environmental attitudes. This trend may be driven by heightened environmental awareness and increased exposure to climate change discussions among younger generations, often through social media and educational initiatives. These results suggest that lifestyle factors and generational attitudes significantly shape behavioral patterns, with younger individuals gravitating toward actions that are both socially encouraged and accessible.

## The Role of Education in Sustainable Actions {#sec-second-point}
Education significantly influences individuals' likelihood of engaging in more substantial climate-friendly behaviors. Our models demonstrate that individuals with higher educational attainment are more likely to adopt behaviors requiring greater financial or logistical investment, such as purchasing electric vehicles or making energy-efficient home improvements. The accuracy of the models predicting these behaviors is relatively lower than for simpler actions, suggesting that education is an important but not sole determinant. This relationship between education and sustainability underscores the importance of knowledge and informed decision-making in supporting climate-friendly actions, particularly those with long-term environmental benefits. The data suggests that higher education increases both environmental awareness and the capacity to implement financially demanding sustainable practices, which are often inaccessible to those with lower educational levels.

## The Influence of Climate Change Knowledge {#sec-third-point}
Self-reported knowledge of climate change also plays a role in the likelihood of adopting sustainable behaviors. Our analysis shows that individuals with greater awareness are more proactive in making environmentally conscious choices, particularly for actions like reducing waste and conserving energy. However, the model accuracy for some knowledge-based behaviors remains moderate, highlighting a gap between awareness and action. This suggests that while raising awareness is critical, it must be paired with policies that reduce structural barriers to adopting more expensive or long-term sustainable actions. For example, financial incentives or subsidies for purchasing energy-efficient appliances could bridge this gap, ensuring individuals are both informed and able to make practical changes.

@@

One significant limitation of this study is its reliance on self-reported survey data, which is inherently prone to biases such as social desirability bias and recall errors. Respondents may overestimate their environmental knowledge or exaggerate their participation in climate-friendly behaviors to present themselves in a favorable light. This introduces uncertainty into the accuracy of the models, as the input data may not fully reflect actual behaviors or knowledge levels. Additionally, self-reported data often lacks the granularity needed to explore complex interactions between socio-economic factors, such as income and household size, which can significantly influence sustainable actions.

Another limitation arises from the use of summary data from the 2018 and 2021 surveys instead of individual-level data. While the summary statistics provided valuable insights into broad demographic trends, they limited the study’s ability to explore more nuanced patterns and interactions. For example, the summary data showed differences in education and age trends between 2018 and 2021, but without individual-level data, it was difficult to determine the direct effect of socio-economic factors on climate action likelihood. Future studies with access to raw, individual-level data could offer more precise estimates and a deeper understanding of these dynamics.

## Challenges with Survey Data
Survey data presents additional challenges, such as non-response bias, where certain demographic groups may be underrepresented, leading to skewed results. Additionally, the framing of survey questions can influence how respondents perceive and answer them, introducing further variability in the data. The inability to verify self-reported behaviors also limits the reliability of the conclusions drawn from such datasets. Despite these challenges, surveys remain a valuable tool for capturing perceptions and self-reported behaviors, but their findings should be interpreted with caution and, when possible, supplemented with objective data sources.

## Future Directions
Future research should aim to identify the specific barriers preventing certain demographic groups from adopting climate-friendly behaviors. This could involve examining factors such as income, access to resources, and perceived obstacles to sustainable practices. Additionally, targeted interventions, such as financial subsidies, public awareness campaigns, or infrastructure improvements, could make sustainable actions more accessible and affordable. Longitudinal studies would be particularly valuable for tracking how awareness of climate change influences sustained behavioral changes over time. Comparing individual-level data across different years could also provide insights into the long-term impact of educational and policy interventions on climate-friendly behaviors.

Lastly, integrating objective data, such as energy consumption records or transportation usage statistics, could validate and enhance the accuracy of survey-based findings. This combined approach would offer a more comprehensive understanding of how demographic and socio-economic factors influence the adoption of sustainable actions, ultimately informing more effective climate policies and interventions.
